# The National Youth Fitness Survey (`nyfs1`) {#NYFS-Study}

The `nyfs1.csv` data file comes from the [2012 National Youth Fitness Survey](http://www.cdc.gov/Nchs/Nnyfs.htm). 

> The NHANES National Youth Fitness Survey (NNYFS) was conducted in 2012 to collect data on physical activity and fitness levels in order to provide an evaluation of the health and fitness of children in the U.S. ages 3 to 15. The NNYFS collected data on physical activity and fitness levels of our youth through interviews and fitness tests.

In the `nyfs1.csv` data file, I'm only providing a tiny portion of the available information. More on the NNYFS (including information I'm not using) is available at the links below.

- [Demographic Information](http://wwwn.cdc.gov/nchs/nhanes/search/nnyfsdata.aspx?Component=Demographics) including a complete [description of all available variables](http://www.cdc.gov/Nchs/Nnyfs/Y_DEMO.htm).
- [Body Measures, part of the general Examination data](http://wwwn.cdc.gov/nchs/nhanes/search/nnyfsdata.aspx?Component=Examination) with complete [variable descriptions](http://www.cdc.gov/Nchs/Nnyfs/Y_BMX.htm)

What I did was merge a few elements from the available demographic information with some elements from the body measures data, reformulated and simplified some variables, and restricted the sample to kids who had a complete set of body measure examinations.

## Looking over the Data Set

To start with, I'll take a look at the `nyfs1` data. One approach is to simply get the size of the set and the names of the available data elements.

```{r nyfs1_read}
## first, we'll import the data into the nyfs1 data frame
nyfs1 <- read.csv("data/nyfs1.csv")

## next we'll turn that data frame into a more useful tibble
nyfs1 <- tbl_df(nyfs1)

## size of the data frame
dim(nyfs1)
```

There are `r dim(nyfs1)[1]` rows (subjects) and `r dim(nyfs1)[2]` columns (variables), by which I mean that there are `r length(nyfs1$subject.id)` kids in the `nyfs1` data frame, and we have `r dim(nyfs1)[2]` pieces of information on each subject. 

So, what do we have, exactly?

```{r nyfs1_tibble}
nyfs1 # this is a tibble, has some nice features in a print-out like this
```

Tibbles are a modern reimagining of the main way in which people have stored data in R, called a data frame. Tibbles were developed to keep what time has proven to be effective, and throwing out what is not. We can obtain the structure of the tibble from the `str` function.

```{r nyfs1_structure}
str(nyfs1)
```

### subject.id

The first variable, `subject.id` is listed by R as an `int` variable, for integer, which means it consists of whole numbers. However, the information provided by this variable is minimal. This is just an identifying code attributable to a given subject of the survey. This is *nominal* data, which will be of little interest down the line. On some occasions, as in this case, the ID numbers are sequential, in the sense that subject 71919 was included in the data base after subject 71918, but this fact isn't particularly interesting here, because the protocol remained unchanged throughout the study.

### sex

The second variable, `sex` is listed as a factor (R uses **factor** to refer to categorical, especially non-numeric information) with two levels,  *Female* and *Male*. You'll note that what is stored in the structure is a series of 1 (referring to the first level - Female) and 2 (Male) values. If we want to know how many people fall in each category, we can build a little table.

```{r nyfs1_sextable}
dplyr::select(nyfs1, sex) %>%
  table()

dplyr::select(nyfs1, sex) %>%
  table() %>%
  addmargins() ## add marginal totals

dplyr::select(nyfs1, sex) %>%
  table() %>%
  prop.table() ## look at the proportions instead
```

Obviously, we don't actually need more than a couple of decimal places for any real purpose.

### age.exam

The third variable, `age.exam` is the age of the child at the time of the examination, measured in years. Note that age is a continuous concept, but the measure used here (number of full years alive) is a common discrete approach to measurement. Age, of course, has a meaningful zero point, so this can be thought of as a ratio variable; a child who is 6 is half as old as one who is 12. We can get a table of the observed values.

```{r nyfs1_ageexamtable}
dplyr::select(nyfs1, age.exam) %>%
  table() %>%
  addmargins()
```

Note that some of the children apparently turned 16 between the time they were initially screened (when they were required to be between 3 and 15 years of age) and the time of the examination. The `sum` listed here is just the total count of all subjects. Since this is a meaningful quantitative variable, we may be interested in a more descriptive summary.

```{r nyfs1_ageexamsummary}
dplyr::select(nyfs1, age.exam) %>%
  summary()
```

These six numbers provide a nice, if incomplete, look at the ages.

* `Min.` = the minimum, or youngest age at the examination was `r min(nyfs1$age.exam)` years old.
* `1st Qu.` = the first quartile (25th percentile) of the ages was `r quantile(nyfs1$age.exam, 0.25)`. This means that 25 percent of the subjects were age `r quantile(nyfs1$age.exam, 0.25)` or less.
* `Median` = the second quartile (50th percentile) of the ages was `r median(nyfs1$age.exam)`. This is often used to describe the center of the data. Half of the subjects were age `r median(nyfs1$age.exam)` or less.
* `3rd Qu.` = the third quartile (75th percentile) of the ages was `r quantile(nyfs1$age.exam, 0.75)`
* `Max.` = the maximum, or oldest age at the examination was `r max(nyfs1$age.exam)` years.

### bmi

The fourth variable, `bmi`, is the body-mass index of the child. The BMI is a person's weight in kilograms divided by his or her height in meters squared. Symbolically, BMI = weight in kg / (height in m)^2^. This is a continuous concept, measured to as many decimal places as you like, and it has a meaningful zero point, so it's a ratio variable.

```{r nyfs1_bmisummary}
dplyr::select(nyfs1, bmi) %>%
  summary()
```

Why would a table of these BMI values not be a great idea, for these data? A hint is that R represents this variable as `num` or numeric in its depiction of the data structure, and this implies that R has some decimal values stored.

```{r nyfs1_bmitable}
dplyr::select(nyfs1, bmi) %>%
  table()
```

### bmi.cat

Our next variable, `bmi.cat`, is a four-category ordinal variable, which divides the sample according to BMI into four groups. The BMI categories use sex-specific 2000 BMI-for-age (in months) growth charts prepared by the Centers for Disease Control for the US. We can get the breakdown from a table of the variable's values. 

```{r nyfs1_bmicattable}
dplyr::select(nyfs1, bmi.cat) %>%
  table() %>%
  addmargins()
```

In terms of percentiles by age and sex from the growth charts, the meanings of the categories are:

* Underweight (BMI < 5th percentile)
* Normal weight (BMI 5th to < 85th percentile)
* Overweight (BMI 85th to < 95th percentile)
* Obese (BMI $\geq$ 95th percentile) 

Note how I've used labels in the `bmi.cat` variable that include a number at the start so that the table results are sorted in a rational way. R sorts tables alphabetically, in general. 

### waist.circ

The sixth variable is `waist.circ`, which is the circumference of the child's waist, in centimeters. Again, this is a numeric variable, so perhaps we'll stick to the simple summary, rather than obtaining a table of observed values.

```{r nyfs1_waist.circsummary}
dplyr::select(nyfs1, waist.circ) %>%
summary()
```

### triceps.skinfold

The seventh and final variable is `triceps.skinfold`, which is measured in millimeters. This is one of several common locations used for the assessment of body fat using skinfold calipers, and is a frequent part of growth assessments in children. Again, this is a numeric variable according to R.

```{r nyfs1$triceps.skinfoldsummary}
dplyr::select(nyfs1, triceps.skinfold) %>%
  summary()
```



## Summarizing the Data Set

The `summary` function can be applied to the whole tibble. For numerical and integer variables, this function produces the five number summary, plus the mean. For categorical (factor) variables, it lists the count for each category.
```{r nyfs1_summary1}
summary(nyfs1)
```

### The Five Number Summary, Quantiles and IQR

The **five number summary** is most famous when used to form a box plot - it's the minimum, 25th percentile, median, 75th percentile and maximum.  Our usual `summary` adds the mean.
```{r nyfs1_summary2}
nyfs1 %>%
    select(bmi) %>%
    summary()
```

As an alternative, we can use the `$` notation to indicate the variable we wish to study inside a data set, and we can use the `fivenum` function to get the five numbers used in developing a box plot.

```{r nyfs1_summary3}
fivenum(nyfs1$bmi)
```

- As mentioned in \@ref(rangeandiqr), the **inter-quartile range**, or IQR, is sometimes used as a competitor for the standard deviation. It's the difference between the 75th percentile and the 25th percentile. The 25th percentile, median, and 75th percentile are referred to as the quartiles of the data set, because, together, they split the data into quarters.

```{r nyfs1_iqr}
IQR(nyfs1$bmi)
```

We can obtain **quantiles** (percentiles) as we like - here, I'm asking for the 1st and 99th
```{r nyfs1_quantiles}
quantile(nyfs1$bmi, probs=c(0.01, 0.99))
```

## Additional Summaries from `favstats`

If we're focusing on a single variable, the `favstats` function in the `mosaic` package can be very helpful. Rather than calling up the entire `mosaic` library here, I'll just specify the function within the library.

```{r nyfs1_favstatsbmi}
mosaic::favstats(nyfs1$bmi)
```

This adds three useful results to the base summary - the standard deviation, the sample size and the number of missing observations.

## The Histogram

As we saw in \@ref(dataviz), obtaining a basic **histogram** of, for example, the BMIs in the `nyfs1` data is pretty straightforward. 

```{r nyfs1_bmihist-fig}
ggplot(data = nyfs1, aes(x = bmi)) +
    geom_histogram(binwidth = 1)
```

### Freedman-Diaconis Rule to select bin width

If we like, we can suggest a particular number of cells for the histogram, instead of accepting the defaults. In this case, we have $n$ = `r length(nyfs1$bmi)` observations. The **Freedman-Diaconis rule** can be helpful here. That rule suggests that we set the bin-width to

\[
h = \frac{2*IQR}{n^{1/3}}
\]

so that the number of bins is equal to the range of the data set (maximum - minimum) divided by $h$. 

For the `bmi` data in the `nyfs1` tibble, we have 

- IQR of `r IQR(nyfs1$bmi)`, $n$ = `r length(nyfs1$bmi)` and range = `r diff(range(nyfs1$bmi))`
- Thus, by the Freedman-Diaconis rule, the optimal binwidth $h$ is `r 2 * IQR(nyfs1$bmi) / length(nyfs1$bmi)^(1/3)`, or, realistically, `r round(2 * IQR(nyfs1$bmi) / length(nyfs1$bmi)^(1/3),0)`. 
- And so the number of bins would be `r diff(range(nyfs1$bmi)) / (2 * IQR(nyfs1$bmi) / length(nyfs1$bmi)^(1/3))`, or, realistically `r round(diff(range(nyfs1$bmi)) / (2 * IQR(nyfs1$bmi) / length(nyfs1$bmi)^(1/3)),0)`.

Here, we'll draw the graph again, using the Freedman-Diaconis rule to identify the number of bins, and also play around a bit with the fill and color of the bars.

```{r nyfs1_bmihist2-fig}
bw <- 2 * IQR(nyfs1$bmi) / length(nyfs1$bmi)^(1/3)
ggplot(data = nyfs1, aes(x = bmi)) +
    geom_histogram(binwidth=bw, color = "white", fill = "black")
```

This is a nice start, but it is by no means a finished graph. 

Let's improve the axis labels, add a title, and fill in the bars with a distinctive blue and use a black outline around each bar. I'll just use 25 bars, because I like how that looks in this case, and optimizing the number of bins is rarely important.

```{r nyfs1_bmihist3-fig}
ggplot(data = nyfs1, aes(x = bmi)) +
    geom_histogram(bins=25, color = "black", fill = "dodgerblue") + 
    labs(title = "Histogram of Body-Mass Index Results in the nyfs1 data",
         x = "Body-Mass Index", y = "# of Patients")
```

## A Note on Colors

The simplest way to specify a color is with its name, enclosed in parentheses. My favorite list of R colors is 
http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf. In  a pinch, you can find it by googling **Colors in R**. You can also type `colors()` in the R console to obtain a list of the names of the same 657 colors.

When using colors to make comparisons, you may be interested in using a scale that has some nice properties. I suggest the `viridis` package to help with this work. The [viridis package vignette](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) describes four color scales (viridis, magma, plasma and inferno) that are designed to be colorful, robust to colorblindness and gray scale printing, and perceptually uniform, which means (as the package authors describe it) that values close to each other have similar-appearing colors and values far away from each other have more different-appearing colors, consistently across the range of values. 

## The Stem-and-Leaf 

We might consider a **stem-and-leaf display** (a John Tukey invention) to show the actual data values while retaining the shape of a histogram. The `scale` parameter can help expand the size of the diagram, so you can see more of the values. Stem and leaf displays are usually used for relatively small samples, perhaps with 10-200 observations, so we'll first take a sample of 150 of the BMI values from the complete set gathered in the `nyfs1` tibble.

```{r nyfs1_sample150_stem-and-leaf}
set.seed(431) # set a seed for the random sampling so we can replicate the results

sampleA <- sample_n(nyfs1, 150, replace = FALSE) # draw a sample of 150 unique rows from nyfs1

stem(sampleA$bmi) # build a stem-and-leaf for those 150 sampled BMI values
```

We can see that the minimum BMI value in this small sample is `r min(sampleA$bmi)` and the maximum BMI value is `r max(sampleA$bmi)`. 

Here's a summary of all variables for these 150 observations.

```{r nyfs1_sample150_summary}
summary(sampleA)
```

If we really wanted to, we could obtain a stem-and-leaf of all of the BMI values in the entire `nyfs1` data. The `scale` parameter lets us see some more of the values.

```{r nyfs1_stemandleafBMI}
stem(nyfs1$bmi, scale = 2)
```

Note that some of the rows extend far beyond what is displayed in the data (as indicated by the `+` sign, followed by a count of the number of unshown data values.)

### A Fancier Stem-and-Leaf Display

We can use the `stem.leaf` function in the `aplpack` package to obtain a fancier version of the stem-and-leaf plot, that identifies outlying values. Below, we display this new version for the random sample of 150 BMI observations we developed earlier.


```{r nyfs1_150_stemandleafBMI_stem.leaf}
aplpack::stem.leaf(sampleA$bmi)
```

We can also produce back-to-back stem and leaf plots to compare, for instance, body-mass index by sex.

```{r nyfs1_150_stemandleafBMI_stem.leaf_backtoback}
samp.F <- filter(sampleA, sex=="Female")
samp.M <- filter(sampleA, sex=="Male")

aplpack::stem.leaf.backback(samp.F$bmi, samp.M$bmi)
```


## The Dot Plot to display a distribution

We can plot the distribution of a single continuous variable using the `dotplot` geom:

```{r nyfs1_bmidotplot-fig}
ggplot(data = nyfs1, aes(x = bmi)) +
    geom_dotplot(dotsize = 0.05, binwidth=1) + 
    scale_y_continuous(NULL, breaks = NULL) + # hides y-axis since it is meaningless
    labs(title = "Dotplot of nyfs1 Body-Mass Index data",
         x = "Body-Mass Index")
```


## The Frequency Polygon

We can plot the distribution of a single continuous variable using the `freqpoly` geom:

```{r nyfs1_bmifreqpoly-fig}
ggplot(data = nyfs1, aes(x = bmi)) +
    geom_freqpoly(binwidth = 1, color = "dodgerblue") + 
    labs(title = "Frequency Polygon of nyfs1 Body-Mass Index data",
         x = "Body-Mass Index", y = "# of Patients")
```


## Plotting the Probability Density Function

We can also produce a density function, which has the effect of smoothing out the bumps in a histogram or frequency polygon, while also changing what is plotted on the y-axis.

```{r nyfs1_bmidensity-fig}
ggplot(data = nyfs1, aes(x = bmi)) +
    geom_density(kernel = "gaussian", color = "dodgerblue") + 
    labs(title = "Density of nyfs1 Body-Mass Index data",
         x = "Body-Mass Index", y = "Probability Density function")
```

So, what's a density function?

- A probability density function is a function of a continuous variable, x, that represents the probability of x falling within a given range. Specifically, the integral over the interval (a,b) of the density function gives the probability that the value of x is within (a,b).
- If you're interested in exploring more on the notion of density functions for continuous (and discrete) random variables, some nice elementary material is available at [Khan Academy](https://www.khanacademy.org/math/statistics-probability/random-variables-stats-library/discrete-and-continuous-random-variables/v/probability-density-functions).


## The Boxplot

Sometimes, it's helpful to picture the five-number summary of the data in such a way as to get a general sense of the distribution. One approach is a **boxplot**, sometimes called a box-and-whisker plot.

```{r nyfs1_boxplot1-fig}
boxplot(nyfs1$bmi, col="yellow", horizontal=T, xlab="Body-Mass Index", 
        main="BMI for 1416 kids in the NYFS")
```

The boxplot is another John Tukey invention. 

- R draws the box (here in yellow) so that its edges of the box fall at the 25^th^ and 75^th^ percentiles of the data, and the thick line inside the box falls at the median (50^th^ percentile). 
- The whiskers then extend out to the largest and smallest values that are not classified by the plot as candidate *outliers*. 
- An outlier is an unusual point, far from the center of a distribution. 
- Note that I've used the `horizontal` option to show this boxplot in this direction. Most comparison boxplots, as we'll see below, are oriented vertically.

The boxplot's **whiskers** that are drawn from the first and third quartiles (i.e. the 25^th^ and 75^th^ percentiles) out to the most extreme points in the data that do not meet the standard of ``candidate outliers.'' An outlier is simply a point that is far away from the center of the data - which may be due to any number of reasons, and generally indicates a need for further investigation. 

Most software, including R, uses a standard proposed by Tukey which describes a ``candidate outlier'' as any point above the ***upper fence*** or below the ***lower fence***. The definitions of the fences are based on the inter-quartile range (IQR).

If IQR = 75th percentile - 25th percentile, then the upper fence is 75th percentile + 1.5 IQR, and the lower fence is 25th percentile - 1.5 IQR.

So for these BMI data,

- the upper fence is located at `r fivenum(nyfs1$bmi)[4]` + 1.5(`r IQR(nyfs1$bmi)`) = `r fivenum(nyfs1$bmi)[4] + 1.5*IQR(nyfs1$bmi)`
- the lower fence is located at `r fivenum(nyfs1$bmi)[2]` - 1.5(`r IQR(nyfs1$bmi)`) = `r fivenum(nyfs1$bmi)[2] - 1.5*IQR(nyfs1$bmi)`

In this case, we see no points identified as outliers in the low part of the distribution, but quite a few identified that way on the high side. This tends to identify about 5\% of the data as a candidate outlier, *if* the data follow a Normal distribution. 

- This plot is indicating clearly that there is some asymmetry (skew) in the data, specifically right skew.
- The standard R uses is to indicate as outliers any points that are more than 1.5 inter-quartile ranges away from the edges of the box. 

The horizontal orientation I've chosen here clarifies the relationship of direction of skew to the plot. A plot like this, with multiple outliers on the right side is indicative of a long right tail in the distribution, and hence, positive or right skew - with the mean being larger than the median. Other indications of skew include having one side of the box being substantially wider than the other, or one side of the whiskers being substantially longer than the other. More on skew later.

### Drawing a Boxplot for One Variable in `ggplot2`

The `ggplot2` library easily handles comparison boxplots for multiple distributions, as we'll see in a moment. However, building a boxplot for a single distribution requires a little trickiness.

```{r nyfs1_boxplot2-fig, fig.height = 3}
ggplot(nyfs1, aes(x = 1, y = bmi)) + 
    geom_boxplot(fill = "yellow") + 
    coord_flip() + 
    labs(title = "Boxplot of BMI for 1416 kids in the NYFS",
         y = "Body Mass Index",
         x = "") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```


## A Simple Comparison Boxplot

Boxplots are most often used for comparison. We can build boxplots using `ggplot2`, as well, and we'll discuss that in detail later. For now, here's a boxplot built to compare the `bmi` results by the child's sex.

```{r nyfs1_boxplot3-fig}
ggplot(nyfs1, aes(x = factor(sex), y = bmi, fill=factor(sex))) +
    geom_boxplot()
```



Let's look at the comparison of observed BMI levels across the four categories in our `bmi.cat` variable, now making use of the `viridis` color scheme. 

```{r nyfs1_boxplot4-fig}
ggplot(nyfs1, aes(x = factor(bmi.cat), y = bmi, fill = factor(bmi.cat))) +
    geom_boxplot() + 
    scale_fill_viridis(discrete=TRUE) + 
    # above line uses viridis palette to identify color choices 
    labs(title = "Observed BMI by BMI Percentile Category, in NYFS1 data")
```

Note that the BMI categories incorporate additional information (in particular the age and sex of the child) beyond the observed BMI, and so the observed BMI levels overlap quite a bit across the four categories. As a graph, that's not bad, but what if we want to improve it further?

Let's turn the boxes in the horizontal direction, and get rid of the perhaps unnecessary `bmi.cat` labels.

```{r nyfs1_boxplot5-fig}
ggplot(nyfs1, aes(x = factor(bmi.cat), y = bmi, fill = factor(bmi.cat))) +
    geom_boxplot() + 
    scale_fill_viridis(discrete=TRUE) + 
    coord_flip() + 
    guides(fill=FALSE) +
    labs(title = "Observed BMI by BMI Percentile Category, in NYFS1 data", x = "")
```

## Using `describe` in the `psych` library

For additional numerical summaries, one option would be to consider using the `describe` function from the `psych` library.

```{r nyfs1_psych_describe_bmi, message=FALSE}
psych::describe(nyfs1$bmi)
```

This package provides, in order, the following...

- `n` = the sample size
- `mean` = the sample mean
- `sd` = the sample standard deviation
- `median` = the median, or 50th percentile
- `trimmed` = mean of the middle 80\% of the data
- `mad` = median absolute deviation
- `min` = minimum value in the sample
- `max` = maximum value in the sample
- `range` = max - min
- `skew` = skewness measure, described below (indicates degree of asymmetry)
- `kurtosis` = kurtosis measure, described below (indicates heaviness of tails, degree of outlier-proneness)
- `se` = standard error of the sample mean = sd / square root of sample size, useful in inference

### The Trimmed Mean

The **trimmed mean** trim value in R indicates proportion of observations to be trimmed from each end of the outcome distribution before the mean is calculated. The `trimmed` value provided by the `psych::describe` package describes what this particular package calls a 20\% trimmed mean (bottom and top 10\% of BMIs are removed before taking the mean - it's the mean of the middle 80\% of the data.) I might call that a 10\% trimmed mean in some settings, but that's just me.

```{r nyfs1_bmi_trimmedmean}
mean(nyfs1$bmi, trim=.1) 
```

### The Median Absolute Deviation

An alternative to the IQR that is fancier, and a bit more robust, is the **median absolute deviation**, which, in large sample sizes, for data that follow a Normal distribution, will be (in expectation) equal to the standard deviation. The MAD is the median of the absolute deviations from the median, multiplied by a constant (1.4826) to yield asymptotically normal consistency.
```{r nyfs1_bmi_mad}
mad(nyfs1$bmi)
```

## Assessing Skew

A relatively common idea is to assess **skewness**, several measures of which (including the one below, sometimes called type 3 skewness, or Pearson's moment coefficient of skewness) are available. Many models assume a Normal distribution, where, among other things, the data are symmetric around the mean. 

Skewness measures asymmetry in the distribution - left skew (mean < median) is indicated by negative skewness values, while right skew (mean > median) is indicated by positive values. The skew value will be near zero for data that follow a Normal distribution. 

### Non-parametric Skew via skew~1~

A simpler measure of skew, sometimes called the **nonparametric skew** and closely related to Pearson's notion of median skewness, falls between -1 and +1 for any distribution. It is just the difference between the mean and the median, divided by the standard deviation. 

- Values greater than +0.2 are sometimes taken to indicate fairly substantial right skew, while values below -0.2 indicate fairly substantial left skew.

```{r nyfs1_skew1_bmi}
(mean(nyfs1$bmi) - median(nyfs1$bmi))/sd(nyfs1$bmi)
```

There is a function in the `Love-boost.R` script called `skew1` that can be used to do these calculations, so long as the variable has no missing data.

```{r useskew1_nyfs1_bmi}
skew1(nyfs1$bmi)
```

[The Wikipedia page on skewness](https://en.wikipedia.org/wiki/Skewness), from which some of this material is derived, provides definitions for several other skewness measures.

## Assessing Kurtosis (Heavy-Tailedness)

Another measure of a distribution's shape that can be found in the `psych` library is the **kurtosis**. Kurtosis is an indicator of whether the distribution is heavy-tailed or light-tailed as compared to a Normal distribution. Positive kurtosis means more of the variance is due to outliers - unusual points far away from the mean relative to what we might expect from a Normally distributed data set with the same standard deviation.

- A Normal distribution will have a kurtosis value near 0, a distribution with similar tail behavior to what we would expect from a Normal is said to be *mesokurtic*
- Higher kurtosis values (meaningfully higher than 0) indicate that, as compared to a Normal distribution, the observed variance is more the result of extreme outliers (i.e. heavy tails) as opposed to being the result of more modest sized deviations from the mean. These heavy-tailed, or outlier prone, distributions are sometimes called *leptokurtic*.
- Kurtosis values meaningfully lower than 0 indicate light-tailed data, with fewer outliers than we'd expect in a Normal distribution. Such distributions are sometimes referred to as *platykurtic*, and include distributions without outliers, like the Uniform distribution.

Here's a table:

Fewer outliers than a Normal | Approximately Normal | More outliers than a Normal
:--------------------------: | :------------------: | :-------------------------:
Light-tailed | "Normalish" | Heavy-tailed
*platykurtic* (kurtosis < 0) | *mesokurtic* (kurtosis = 0) | *leptokurtic* (kurtosis > 0)

```{r nyfs1_bmi_kurtosis}
psych::kurtosi(nyfs1$bmi)
```


### The Standard Error of the Sample Mean

The **standard error** of the sample mean, which is the standard deviation divided by the square root of the sample size:

```{r nyfs1_bmi_se}
sd(nyfs1$bmi)/sqrt(length(nyfs1$bmi))
```

## The `describe` function in the `Hmisc` library

The `Hmisc` library has lots of useful functions. It's named for its main developer, Frank Harrell. The `describe` function in `Hmisc` knows enough to separate numerical from categorical variables, and give you separate (and detailed) summaries for each.

- For a categorical variable, it provides counts of total observations (n), the number of missing values, and the number of unique categories, along with counts and percentages falling in each category.
- For a numerical variable, it provides:
+ counts of total observations (n), the number of missing values, and the number of unique values
+ an Info value for the data, which indicates how continuous the variable is (a score of 1 is generally indicative of a completely continuous variable with no ties, while scores near 0 indicate lots of ties, and very few unique values)
+ the sample Mean
+ many sample percentiles (quantiles) of the data, specifically (5, 10, 25, 50, 75, 90, 95, 99)
+ either a complete table of all observed values, with counts and percentages (if there are a modest number of unique values), or
+ a table of the five smallest and five largest values in the data set, which is useful for range checking

```{r numerical summary from Hmisc}
Hmisc::describe(nyfs1)
```

More on the `Info` value in Hmisc::describe is [available here](https://cran.r-project.org/web/packages/Hmisc/Hmisc.pdf#page=67)

## `xda` from GitHub for numerical summaries for exploratory data analysis

```{r using xda 1}
## next two commands needed if xda is not already installed

library(devtools)
install_github("ujjwalkarn/xda")

xda::numSummary(nyfs1)
```

Most of the elements of this `numSummary` should be familiar. Some new pieces include:

- `nunique` = number of unique values
- `nzeroes` = number of zeroes
- `noutlier` = number of outliers (using a standard that isn't entirely transparent to me)
- `miss` = number of rows with missing value
- `miss%` = percentage of total rows with missing values ((miss/n)*100)
- `5%` = 5th percentile value of that variable (value below which 5 percent of the observations may be found)

```{r using xda 2}
xda::charSummary(nyfs1)
```

The `top5levels:count` provides the top 5 unique values for each variable, sorted by their counts.

## What Summaries to Report

It is usually helpful to focus on the shape, center and spread of a distribution. Bock, Velleman and DeVeaux provide some useful advice:
    
- If the data are skewed, report the median and IQR (or the three middle quantiles). You may want to include the mean and standard deviation, but you should point out why the mean and median differ. The fact that the mean and median do not agree is a sign that the distribution may be skewed. A histogram will help you make that point.
- If the data are symmetric, report the mean and standard deviation, and possibly the median and IQR as well.
- If there are clear outliers and you are reporting the mean and standard deviation, report them with the outliers present and with the outliers removed. The differences may be revealing. The median and IQR are not likely to be seriously affected by outliers.
