# The Western Collaborative Group Study

## The Western Collaborative Group Study (`wcgs`) data set

@Vittinghoff explore data from the Western Collaborative Group Study (WCGS) in great detail^[For more on the WCGS, you might look at http://www.epi.umn.edu/cvdepi/study-synopsis/western-collaborative-group-study/]. We'll touch lightly on some key issues in this Chapter.

```{r tibble_wcgs_data}
wcgs <- read.csv("data/wcgs.csv") %>% tbl_df

wcgs
```

Here, we have `r dim(wcgs)[1]` rows (subjects) and `r dim(wcgs)[2]` columns (variables). 

### Structure of `wcgs`

We can specify the (sometimes terrible) variable names, through the `names` function, or we can add other elements of the structure, so that we can identify elements of particular interest.

```{r wcgs_structure}
str(wcgs)
```

### Codebook for `wcgs`

This table was lovingly hand-crafted, and involved a lot of typing. We'll look for better ways in 432.

Name    | Stored As  | Type         | Details (units, levels, etc.)
-------:|:----------:|:------------:|------------------------------------
`id`      | integer    | (nominal)  | ID #, nominal and uninteresting
`age`     | integer    | quantitative | age, in years - no decimal places
`agec`    | factor (5) | (ordinal) | age: 35-40, 41-45, 46-50, 51-55, 56-60
`height`  | integer    | quantitative | height, in inches
`weight`  | integer    | quantitative | weight, in pounds
`lnwght`  | number     | quantitative | natural logarithm of weight
`wghtcat` | factor (4) | (ordinal) | wt: < 140, 140-170, 170-200, > 200
`bmi`     | number     | quantitative | body-mass index: 
| | | | 703 * weight in lb / (height in in)^2^
`sbp`     | integer    | quantitative | systolic blood pressure, in mm Hg
`lnsbp`   | number     | quantitative | natural logarithm of sbp
`dbp`     | integer    | quantitative | diastolic blood pressure, mm Hg
`chol`    | integer    | quantitative | total cholesterol, mg/dL
`behpat`  | factor (4) | (nominal) | behavioral pattern: A1, A2, B3 or B4
`dibpat`  | factor (2) | (binary)  | behavioral pattern: A or B
`smoke`   | factor (2) | (binary)  | cigarette smoker: Yes or No
`ncigs`   | integer | quantitative | number of cigarettes smoked per day
`arcus`   | integer | (nominal) | arcus senilis present (1) or absent (0)
`chd69`   | factor (2) | (binary) | CHD event: Yes or No
`typchd69`| integer | (4 levels) | event: 0 = no CHD, 1 = MI or SD,
| | | | 2 = silent MI, 3 = angina
`time169` | integer | quantitative | follow-up time in days
`t1`      | number  | quantitative | heavy-tailed (random draws)
`uni`     | number  | quantitative | light-tailed (random draws)

### Quick Summary

```{r summary_wcgs_simple}
summary(wcgs)
```

For a more detailed description, we might consider `Hmisc::describe`, `psych::describe`, `mosaic::favstats`, etc.

## Are the SBPs Normally Distributed?

Consider the question of whether the distribution of the systolic blood pressure results is well-approximated by the Normal.

```{r wcgs_sbphistwithnormal-fig}
ggplot(wcgs, aes(x = sbp)) + 
    geom_histogram(aes(y = ..density..), 
                   bins = 30, fill = "aquamarine", col="blue") +
    stat_function(fun = dnorm, lwd = 1.5, col = "darkblue",
                  args = list(mean = mean(wcgs$sbp), sd = sd(wcgs$sbp))) +
    annotate("text", x = 200, y = 0.01, col = "darkblue",
             label = paste("Mean = ", round(mean(wcgs$sbp),1),
                           ", SD = ", round(sd(wcgs$sbp),1))) +
    labs(title = "Systolic BP for `wcgs` subjects",
         caption = "Histogram with Normal Density superimposed")
```

Since the data contain both `sbp` and `lnsbp` (its natural logarithm), let's compare them. Note that in preparing the graph, we'll need to change the location for the text annotation.

```{r wcgs_lnsbphistwithnormal-fig}
ggplot(wcgs, aes(x = lnsbp)) + 
    geom_histogram(aes(y = ..density..), 
                   bins = 30, fill = "aquamarine", col="blue") +
    stat_function(fun = dnorm, lwd = 1.5, col = "darkblue",
                  args = list(mean = mean(wcgs$lnsbp), 
                              sd = sd(wcgs$lnsbp))) +
    annotate("text", x = 5.2, y = 2, col = "darkblue",
             label = paste("Mean = ", round(mean(wcgs$lnsbp),1),
                           ", SD = ", round(sd(wcgs$lnsbp),1))) +
    labs(title = "Natural Logarithm of `wcgs` Systolic BP",
         caption = "Histogram with Normal Density superimposed")
```

We can also look at Normal Q-Q plots, for instance...

```{r wcgs_qqplotsforsbpandlnsbpfromggplot-fig}
p1 <- ggplot(wcgs, aes(sample = sbp)) +
    geom_point(stat="qq", color = "orchid") + 
    labs(y = "Ordered SBP", title = "sbp in wcgs")

p2 <- ggplot(wcgs, aes(sample = lnsbp)) +
    geom_point(stat="qq", color = "orange") + 
    labs(y = "Ordered ln(sbp)", title = "ln(sbp) in wcgs")

gridExtra::grid.arrange(p1, p2, ncol=2, top ="Normal Q-Q plots of SBP and ln(SBP) in wcgs")
```

There's at best a small improvement from `sbp` to `lnsbp` in terms of approximation by a Normal distribution.

## Describing Outlying Values with Z Scores

It looks like there's an outlier (or a series of them) in the SBP data.

```{r wcgs_boxplotofsbp_alone}
ggplot(wcgs, aes(x = 1, y = sbp)) +
    geom_boxplot(fill = "royalblue") +
    labs(title = "Boxplot of SBP in `wcgs` data",
         y = "Systolic Blood Pressure (mm Hg)", 
         x = "") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank()) +
    coord_flip() 
```


```{r wcgs_sbp_Hmisc_describe}
Hmisc::describe(wcgs$sbp)
```

The maximum value here is `r max(wcgs$sbp)`, and is clearly the most extreme value in the data set. One way to gauge this is to describe that observation's **Z score**, the number of standard deviations away from the mean that the observation falls. Here, the maximum value, `r max(wcgs$sbp)` is `r round((max(wcgs$sbp)-mean(wcgs$sbp))/sd(wcgs$sbp), 2)` standard deviations above the mean, and thus has a Z score of `r round((max(wcgs$sbp)-mean(wcgs$sbp))/sd(wcgs$sbp), 1)`.

A negative Z score would indicate a point below the mean, while a positive Z score indicates, as we've seen, a point above the mean. The minimum systolic blood pressure, `r min(wcgs$sbp)` is `r round((mean(wcgs$sbp)-min(wcgs$sbp))/sd(wcgs$sbp), 2)` standard deviations *below* the mean, so it has a Z score of `r round((min(wcgs$sbp)-mean(wcgs$sbp))/sd(wcgs$sbp), 1)`. 

Recall that the Empirical Rule suggests that if a variable follows a Normal distribution, it would have approximately 95% of its observations falling inside a Z score of (-2, 2), and 99.74% falling inside a Z score range of (-3, 3). Do the systolic blood pressures appear Normally distributed?

## Does Weight Category Relate to SBP?

The data are collected into four groups based on the subject's weight (in pounds).

```{r wcgssbpbyweightbox-fig}
ggplot(wcgs, aes(x = wghtcat, y = sbp, fill = wghtcat)) +
    geom_boxplot(notch = TRUE) +
    scale_fill_viridis(discrete=TRUE) +
    guides(fill = FALSE) + 
    labs(title = "Boxplot of Systolic BP by Weight Category in WCGS", 
         x = "Weight Category", y = "Systolic Blood Pressure")
```

## Re-Leveling a Factor

Well, that's not so good. We really want those weight categories (the *levels*) to be ordered more sensibly. 

```{r levels of wghtcat}
table(wcgs$wghtcat)
```

Like all *factor* variables in R, the categories are specified as levels.

```{r wcgs_levels_of_wghtcat}
levels(wcgs$wghtcat)
```

We want to change the order of the levels in a new version of this factor variable so they make sense. There are multiple ways to do this, but I prefer the `fct_relevel` function from the `forcats` package. Which order is more appropriate?

```{r revise order of wghtcat in new variable}
table(fct_relevel(wcgs$wghtcat, "< 140", "140-170", "170-200", "> 200"), wcgs$wghtcat)
```

I'll add a new variable to the `wcgs` data called `weight_f` that relevels the `wghtcat` data.

```{r create weight_f}
wcgs <- wcgs %>%
    mutate(weight_f = fct_relevel(wghtcat, "< 140", "140-170", "170-200", "> 200"))

table(wcgs$weight_f)
```

For more on the `forcats` package, check out @R4DS, especially the Section on Factors.

### SBP by Weight Category

```{r wcgssbpbyweightboxrevised-fig}
ggplot(wcgs, aes(x = weight_f, y = sbp, fill = weight_f)) +
    geom_boxplot(notch = TRUE) +
    scale_fill_viridis(discrete=TRUE) +
    guides(fill = FALSE) +
    labs(title = "Systolic Blood Pressure by Reordered Weight Category in WCGS", 
         x = "Weight Category", y = "Systolic Blood Pressure")
```

We might see some details well with a **joy plot**, too.

```{r wcgs_sbp_by_weight_f_joyplot}
wcgs %>%
    ggplot(aes(x = sbp, y = weight_f, fill = weight_f, height = ..density..)) +
    ggjoy::geom_joy(scale = 2) +
    scale_fill_viridis(discrete = TRUE) +
    guides(fill = FALSE) +
    labs(title = "Joy Plot of SBP by Weight Category (wcgs)",
         x = "Systolic Blood Pressure",
         y = "Weight Category") +
    theme_bw()
```

As the plots suggest, patients in the heavier groups generally had higher systolic blood pressures.

```{r wcgs_sbp_by_weight_f_summary}
by(wcgs$sbp, wcgs$weight_f, mosaic::favstats)
```


## Are Weight and SBP Linked?

Let's build a scatter plot of SBP (Outcome) by Weight (Predictor), rather than breaking down into categories.

```{r wcgs_scatter_sbpbyweight-fig}
ggplot(wcgs, aes(x = weight, y = sbp)) +
    geom_point(size=3, shape=1, color="forestgreen") + ## default size = 2
    stat_smooth(method=lm, color="red") + ## add se=FALSE to hide conf. interval
    stat_smooth(method=loess, se=FALSE, color="blue") +
    ggtitle("SBP vs. Weight in 3,154 WCGS Subjects") + 
    theme_bw()
```

- The mass of the data is hidden from us - showing `r length(wcgs$sbp)` points in one plot can produce little more than a blur where there are lots of points on top of each other.
- Here the least squares regression line (in red), and loess scatterplot smoother, (in blue) can help.

The relationship between systolic blood pressure and weight appears to be very close to linear, but of course there is considerable scatter around that generally linear relationship. It turns out that the Pearson correlation of these two variables is `r signif(cor(wcgs$sbp, wcgs$weight),3)`.

## SBP and Weight by Arcus Senilis groups?

An issue of interest to us will be to assess whether the SBP-Weight relationship we see above is similar among subjects who have arcus senilis and those who do not. 

> Arcus senilis is an old age syndrome where there is a white, grey, or blue opaque ring in the corneal margin (peripheral corneal opacity), or white ring in front of the periphery of the iris. It is present at birth but then fades; however, it is quite commonly present in the elderly. It can also appear earlier in life as a result of hypercholesterolemia.
> 
> Wikipedia article on Arcus Senilis, retrieved 2017-08-15

Let's start with a quick look at the `arcus` data.

```{r wcgs_summary_arcus1}
wcgs %>%
    select(arcus) %>%
    summary()
```

We have `r sum(is.na(wcgs$arcus))` missing values, so we probably want to do something about that before plotting the data, and we may also want to create a factor variable with more meaningful labels than 1 (which means yes, arcus senilis is present) and 0 (which means no, it isn't.) We'll use the 

```{r wcgs_factor_create_arcus_f}
wcgs <- wcgs %>%
    mutate(arcus_f = fct_recode(factor(arcus),
                                "Arcus senilis" = "1",
                                "No arcus senilis" = "0"),
           arcus_f = fct_relevel(arcus_f, "Arcus senilis"))

table(wcgs$arcus_f, wcgs$arcus, useNA = "ifany")
```

Let's build a version of the `wcgs` data that eliminates all missing data in the variables of immediate interest, and then plot the SBP-weight relationship in groups of patients with and without arcus senilis.

```{r wcgs_scatter_sbpbyweightbyarcus_facets-fig}
wcgs %>%
    filter(complete.cases(arcus_f, sbp, weight)) %>%
    ggplot(aes(x = weight, y = sbp, group = arcus_f)) +
    geom_point(size=3, shape = 1) + 
    stat_smooth(method=lm, color="red") +
    stat_smooth(method=loess, se=FALSE, color="blue") +
    labs(title = "SBP vs. Weight by Arcus Senilis status",
         caption = "3,152 Western Collaborative Group Study subjects with known arcus senilis status") + 
    facet_wrap(~ arcus_f) +
    theme_bw()
```

## Linear Model for SBP-Weight Relationship: subjects without Arcus Senilis

```{r linear model summary-no arcus}
model.noarcus <- 
    lm(sbp ~ weight, data = filter(wcgs, arcus == 0))

summary(model.noarcus)
```

The linear model for the `r table(wcgs$arcus)["0"]` patients without Arcus Senilis has R^2^ = `r 100*signif(summary(model.noarcus)$r.square,3)`\%.

- The regression equation is 95.92 - 0.19 weight, for those patients without Arcus Senilis.

## Linear Model for SBP-Weight Relationship: subjects with Arcus Senilis

```{r linear model summary-with arcus}
model.witharcus <- 
    lm(sbp ~ weight, data = filter(wcgs, arcus == 1))

summary(model.witharcus)
```

The linear model for the `r table(wcgs$arcus)["1"]` patients with Arcus Senilis has R^2^ = `r 100*signif(summary(model.witharcus)$r.square,3)`\%.

- The regression equation is 101.88 - 0.163 weight, for those patients with Arcus Senilis.

## Including Arcus Status in the model

```{r model3_with_arcus_included}
model3 <- lm(sbp ~ weight * arcus, data = filter(wcgs, !is.na(arcus)))

summary(model3)
```

The actual regression equation in this setting includes both weight, and an indicator variable (1 = yes, 0 = no) for arcus senilis status, and the product of weight and that 1/0 indicator.

- Note the use of the product term `weight*arcus` in the setup of the model to allow both the slope of weight and the intercept term in the model to change depending on arcus senilis status.
    + For a patient who has arcus, the regression equation is SBP = 95.92 - 0.19 weight + 5.96 (1) - 0.028 weight (1) = 101.88 + 0.162 weight.
    + For a patient without arcus senilis, the regression equation is SBP = 95.92 - 0.19 weight + 5.96 (0) - 0.028 weight (0) = 95.92 - 0.19 weight. 

The linear model including the interaction of weight and arcus to predict sbp for the `r sum(!is.na(wcgs$arcus))` patients with known Arcus Senilis status has R^2^ = `r 100*signif(summary(model3)$r.square,3)`\%.


## Predictions from these Linear Models

What is our predicted SBP for a subject weighing 175 pounds?
    
How does that change if our subject weighs 200 pounds?
    
Recall that

- *Without* Arcus Senilis, linear model for SBP = `r round(coef(model.noarcus)[1],1)` + `r round(coef(model.noarcus)[2],2)` x weight
- *With* Arcus Senilis, linear model for SBP = `r round(coef(model.witharcus)[1],1)` + `r round(coef(model.witharcus)[2],2)` x weight

So the predictions for a 175 pound subject are:
- `r round(coef(model.noarcus)[1],1)` + `r round(coef(model.noarcus)[2],2)` x 175 = `r round(coef(model.noarcus)[1] + coef(model.noarcus)[2]*175,0)` mm Hg without Arcus Senilis, and 
- `r round(coef(model.witharcus)[1],1)` + `r round(coef(model.witharcus)[2],2)` x 175 = `r round(coef(model.witharcus)[1] + coef(model.witharcus)[2]*175,0)` mm Hg with Arcus Senilis.

And thus, the predictions for a 200 pound subject are:
- `r round(coef(model.noarcus)[1],1)` + `r round(coef(model.noarcus)[2],2)` x 200 = `r round(coef(model.noarcus)[1] + coef(model.noarcus)[2]*200,1)` mm Hg without Arcus Senilis, and 
- `r round(coef(model.witharcus)[1],1)` + `r round(coef(model.witharcus)[2],2)` x 200 = `r round(coef(model.witharcus)[1] + coef(model.witharcus)[2]*200,1)` mm Hg with Arcus Senilis.

```{r clean up after wcgs}
rm(model.noarcus, model.witharcus)
```

## Scatterplots with Facets Across a Categorical Variable

We can use facets in `ggplot2` to show scatterplots across the levels of a categorical variable, like `behpat`.

```{r wcgsweightandsbpbybehavior-fig}
ggplot(wcgs, aes(x = weight, y = sbp, col = behpat)) +
    geom_point() +
    facet_wrap(~ behpat) +
    geom_smooth(method = "lm", se = FALSE, col = "black") +
    guides(color = FALSE) +
    theme(strip.text = element_text(face="bold", size=rel(1.25), color="white"),
          strip.background = element_rect(fill="royalblue")) +
    labs(title = "Scatterplots of SBP vs. Weight within Behavior Pattern")
```

## Scatterplot and Correlation Matrices

A **scatterplot matrix** can be very helpful in understanding relationships between multiple variables simultaneously. There are several ways to build such a thing, including the `pairs` function...

```{r scattermata-fig, fig.height=6}
pairs (~ sbp + age + weight + height, data=wcgs, main="Simple Scatterplot Matrix")
```

### Using the car library

Or, we can use the scatterplotMatrix function from the `car` library, which adds some detail and fitting to the plots, and places density estimates (with rug plots) on the diagonals.

```{r scattermatcres-fig}
car::scatterplotMatrix(~ sbp + age + weight + height, data=wcgs, 
main="Scatterplot Matrix via car")
```

### Displaying a Correlation Matrix

```{r dplyr for correlation matrix}
wcgs %>%
    dplyr::select(sbp, age, weight, height) %>%
    cor() %>% # obtain correlation coefficients for this subgroup
    signif(., 3) # round them off to three significant figures before printing
```

### Augmented Scatterplot Matrix

Dr. Love's favorite way to augment a scatterplot matrix adds LOWESS smoothed lines in the upper panel, and correlations in the lower panel, with histograms down the diagonal. To do this, I revised two functions in the Love-boost script (these modifications come from [Chang's R Graphics Cookbook](https://sites.google.com/a/case.edu/love-431/home/texts-not-written-by-dr-love)), called `panel.hist` and `panel.cor`.

```{r wcgsscattermatrixaugmented-fig}
# requires Love-boost.R is sourced

pairs (~ sbp + age + weight + height, data=wcgs, 
       main="Augmented Scatterplot Matrix", 
       upper.panel = panel.smooth,
       diag.panel = panel.hist,
       lower.panel = panel.cor)
```

### Using the `GGally` package

The `ggplot2` system doesn't have a built-in scatterplot system. There are some nice add-ins in the world, though. One option I sort of like is in the `GGally` package, which can produce both correlation matrices and scatterplot matrices.

```{r GGally correlation matrix}
GGally::ggcorr(select(wcgs, sbp, age, weight, height), name = "Pearson r", label = TRUE)
```

The `ggpairs` function provides a density plot on each diagonal, Pearson correlations on the upper right and scatterplots on the lower left of the matrix.

```{r ggallywcgs-fig}
GGally::ggpairs(select(wcgs, sbp, age, weight, height), title = "Scatterplot Matrix via ggpairs")
```


