# Confidence Intervals from Two Independent Samples of Quantitative Data {#CI-Indep-Samples}

Here, we'll consider the problem of estimating a confidence interval to describe the difference in population means (or medians) based on a comparison of two samples of quantitative data, gathered using an independent samples design. Specifically, we'll use as our example the randomized controlled trial of Ibuprofen in Sepsis patients, as described in Section \@ref(Sepsis_RCT). 

In that trial, 300 patients meeting specific criteria (including elevated temperature) for a diagnosis of sepsis were randomly assigned to either the Ibuprofen group (150 patients) and 150 to the Placebo group. Group information (our exposure) is contained in the `treat` variable. The key outcome of interest to us was `temp_drop`, the change in body temperature (in $^{\circ}$C) from baseline to 2 hours later, so that positive numbers indicate drops in temperature (a good outcome.)

```{r sepsis-tibble-listing-for-ci-section}
sepsis
```

## t-based CI for population mean difference $mu_1 - \mu_2$ from Independent Samples

### The Welch t procedure

The default confidence interval based on the t test for independent samples in R uses something called the Welch test, in which the two populations being compared are not assumed to have the same variance. Each population is assumed to follow a Normal distribution. 

```{r sepsis-ci-t-welch-1}
t.test(sepsis$temp_drop ~ sepsis$treat, conf.level = 0.90, alt = "two.sided")
```

### The Pooled t procedure

The most commonly used t-procedure for building a confidence interval assumes not only that each of the two populations being compared follows a Normal distribution, but also that they have the same population variance. This is the pooled t-test, and it is what people usually mean when they describe a two-sample t test.

```{r sepsis-ci-t-pooled-1}
t.test(sepsis$temp_drop ~ sepsis$treat, conf.level = 0.90, alt = "two.sided", var.equal = TRUE)
```

### Using linear regression to obtain a pooled t confidence interval

A linear regression model, using the same outcome and predictor (group) as the pooled t procedure, produces the same confidence interval, again, under the assumption that the two populations we are comparing follow a Normal distribution with the same (population) variance.

```{r}
model1 <- lm(temp_drop ~ treat, data = sepsis)
model1
confint(model1, level = 0.90)
```

We see that our point estimate from the linear regression model is that the difference in `temp_drop` is `r coef(model1)[2]`, where Ibuprofen subjects have higher `temp_drop` values than do Placebo subjects, and that the 90% confidence interval for this difference ranges from `r confint(model1, level = 0.90)[2]` to `r confint(model1, level = 0.90)[4]`.

We can obtain a t-based confidence interval for each of the parameter estimates in a linear model directly using `confint`. Linear models usually summarize only the estimate and standard error. Remember that a reasonable approximation in large samples to a 95% confidence interval for a regression estimate (slope or intercept) can be obtained from estimate $\pm$ 2 * standard error.

```{r}
summary(model1)
```

So, in the case of the `treatPlacebo` estimate, we can obtain an approximate 95% confidence interval with `r round(coef(summary(model1))["treatPlacebo",1],3)` $\pm$ 2 x `r round(coef(summary(model1))["treatPlacebo",2],3)` or (`r round(coef(summary(model1))["treatPlacebo",1] - 2*coef(summary(model1))["treatPlacebo",2],3)`, `r round(coef(summary(model1))["treatPlacebo",1]+2*coef(summary(model1))["treatPlacebo",2],3)`). Compare this to the 95% confidence interval available from the model directly, shown below, and you'll see only a small difference.

```{r}
confint(model1, level = 0.95)
```

## Bootstrap CI for $mu_1 - \mu_2$ from Independent Samples

The `bootdif` function contained in the `Love-boost.R` script, that we will use in this setting is a slightly edited version of the function at http://biostat.mc.vanderbilt.edu/wiki/Main/BootstrapMeansSoftware. Note that this approach uses a comma to separate the outcome variable (here, `temp_drop`) from the variable identifying the exposure groups (here, `treat`).

```{r sepsis-ci-bootstrap-1}
set.seed(431212)
bootdif(sepsis$temp_drop, sepsis$treat, conf.level = 0.90)
```


## Wilcoxon Rank Sum-based CI from Independent Samples

As in the one-sample case, a rank-based alternative attributed to Wilcoxon (and sometimes to Mann and Whitney) provides a two-sample comparison of the pseudomedians in the two `treat` groups in terms of `temp_drop`. This is called a **rank sum** test, rather than the **signed rank** test for a single sample. Here's the resulting 90% confidence interval.

```{r sepsis-ci-wilcoxon-1}
wilcox.test(sepsis$temp_drop ~ sepsis$treat, 
            conf.int = TRUE, conf.level = 0.90, 
            alt = "two.sided")
```

## Using the `tidy` function from `broom` for t and Wilcoxon procedures

The `tidy` function is again available to us in dealing with a t-test or Wilcoxon rank sum test.

```{r}
broom::tidy(t.test(sepsis$temp_drop ~ sepsis$treat, 
                   conf.level = 0.90, 
                   alt = "two.sided"))

broom::tidy(wilcox.test(sepsis$temp_drop ~ sepsis$treat, 
                        conf.int = TRUE, 
                        conf.level = 0.90, 
                        alt = "two.sided"))
```

We can also use `broom` functions to place the elements of the linear model `model1` into a tidy data frame. This provides the estimate of the Placebo-Ibuprofen difference, and its standard error, which we could use to formulate a confidence interval.

```{r}
broom::tidy(model1)
rm(model1)
```
