# Larger Contingency Tables - Testing for Independence

What will we do with tables describing data from more than two categories at a time, returning to the notion of independent (rather than paired or matched) samples? The chi-square tests we have already seen in our `twobytwo` table output will extend nicely to this scenario, especially the Pearson $\chi^2$ (asymptotic) test.

## A 2x3 Table: Comparing Response to Active vs. Placebo

The table below, containing 2 rows and 3 columns of data (ignoring the marginal totals) specifies the number of patients who show *complete*, *partial*, or *no response* after treatment with either **active** medication or a **placebo**.

Group | None | Partial | Complete
-----:| :---:| :----: | :-----:
Active | 16 | 26 | 29
Placebo | 24 | 26 | 18

Is there a statistically significant association here? That is to say, is there a statistically significant difference between the treatment groups in the distribution of responses?

### Getting the Table into R

To answer this, we'll have to get the data from this contingency table into a matrix in R. Here's one approach...

```{r table2x3 a}
T1 <- matrix(c(16,26,29,24,26,18), ncol=3, nrow=2, byrow=TRUE)
rownames(T1) <- c("Active", "Placebo")
colnames(T1) <- c("None", "Partial", "Complete")

T1
```

### Manipulating the Table's presentation

We can add margins to the matrix to get a table including row and column totals.
```{r table2x3 b}
addmargins(T1)
```

Instead of the counts, we can tabulate the proportion of all patients within each cell.
```{r table2x3 c}
prop.table(T1)
```

Or, we can tabulate the probabilities, rather than the proportions, after some rounding.
```{r table2x3 d}
round(100*prop.table(T1),1)
```

R can also plot the percentages conditional on the rows...
```{r table2x3 e}
round(100*prop.table(T1,1),1)
```

The 40.8 in Active/Complete means that of the Active medication patients, 40.8% had a complete response.

R can also plot the percentages conditional on the columns...

```{r table2x3 g}
round(100*prop.table(T1,2),1)
```

If we add the row of column totals for these percentages as shown below, it clarifies that the 61.7 in Active/Complete here means that of the patients with a Complete response, 61.7% were on the active medication.

```{r table2x3 h}
addmargins(round(100*prop.table(T1,2),1),1)
```

## Getting the Chi-Square Test Results

Now, to actually obtain a *p* value and perform the significance test with H~0~: rows and columns are independent vs. H~A~: rows and columns are associated, we simply run a Pearson chi-square test on T1 ...

```{r table2x3 i}
chisq.test(T1)
```

Thanks to a p-value of about 0.13 (using the Pearson $\chi^2$ test) our conclusion would be to retain the null hypothesis of independence in this setting.

We could have run a Fisher's exact test, too, if we needed it. 

```{r table2x3 j}
fisher.test(T1)
rm(T1)
```

The Fisher exact test *p* value is also 0.13. Either way, there is insufficient evidence to conclude that there is a (true) difference in the distributions of responses.

## Getting a 2x3 Table into R using a .csv file

Suppose we have a table like this available, and want to compute the Pearson and Fisher $\chi^2$ test results in R, without having to set up the whole matrix piece discussed above? 

Group | None | Partial | Complete
-----:| :---:| :----: | :-----:
Active | 16 | 26 | 29
Placebo | 24 | 26 | 18

We could do so by building a .csv file (a spreadsheet) containing the table above. In particular, I built a file called `active2x3.csv`, available on the course website. It is simply the table below.

```
Group   None  Partial	Complete
Active	  16	     26	      29
Placebo	  24	     26	      18
```

When we pull this .csv file into R, it emerges as a data frame and deliberately NOT as a tibble.

```{r active2x3}
active2x3
```

## Turning the Data Frame into a Table That R Recognizes

We need to turn this data frame into something that R can recognize as a contingency table. The steps to do this are:

1. Establish row names from the Group variable.
2. Delete the Group variable, retaining only the variables containing counts.
3. Convert the data frame into a matrix.

Specifically in this case, the steps are:

```{r active2x3 into table}
rownames(active2x3) <- active2x3$Group
active2x3$Group <- NULL
tab2x3 <- as.matrix(active2x3)
tab2x3
```

And this resulting `tab2x3` object can be treated as the matrix was previously, using `addmargins` or `chisq.test` or `prop.table` or whatever. For instance,

```{r tab2x3 fisher}
fisher.test(tab2x3)
```

## Collapsing Levels / Categories in a Factor

Returning to the `survey1` data, let's build a table of the relationship between response to the `sex` and favorite `color` questions.

```{r survey1 colors a}
table(survey1$sex, survey1$color)
chisq.test(table(survey1$sex, survey1$color))
```

Note the warning message. With all those small cell counts, and particular, so many zeros, this might not be so useful. 

We might reasonably consider collapsing the `colors` data into four new categories:

- Blue -- in which I will include the old aqua, aquamarine, blue, dark blue and sapphire; 
- Green, which will include the old chartreuse and green, 
- Red - which will include the old orange, pink and red, and 
- Other – which will include all other colors, including purple and violet

One way to do this sort of work uses the `plyr` library in R, and the `revalue` function in particular. I learned about this in Chapter 15 of Chung's R Graphics Cookbook, which is invaluable.

```{r survey1 colors b}
survey1$color.new <- factor(survey1$color)
levels(survey1$color.new)
survey1$color.new <- dplyr::recode(survey1$color.new, 
    aqua = "Blue", aquamarine="Blue", black = "Other",
    blue = "Blue", brown = "Other", chartreuse = "Green", 
    'dark blue' = "Blue", depends = "Other", gray = "Other",
    green = "Green", grey ="Other", 'light blue' = "Blue",
    navy = "Blue", none = "Other", orange = "Red", 
    pink = "Red", purple = "Other", red = "Red", 
    'royal blue' = "Blue", sapphire = "Blue", silver = "Other",
    teal = "Blue", violet = "Other", white = "Other", yellow = "Other")
```

So, what I've done here is create a new color variable that assigns the original colors into the four categories: Blue, Green, Red and Other that I defined above. Let's run a sanity check on our recoding, then look at the relationship between sex and this new four-category variable, in a 2x4 table...

```{r survey1 colors c}
table(survey1$color, survey1$color.new)  ## sanity check

table(survey1$sex, survey1$color.new)
```

To neaten this output up, we might want to have Other show up last in the `color.new` group.

```{r survey1 colors d}
survey1$color.new2 <- factor(survey1$color.new, 
       levels=c("Blue", "Green", "Red", "Other"))
table(survey1$sex, survey1$color.new2)
```

Now, we run the new Pearson $\chi^2$ test, and conclude that there is no evidence of statistically significant association (at the 5% significance level) between sex and these collapsed favorite color categories.

```{r survey1 colors e}
chisq.test(table(survey1$sex, survey1$color.new2))
```

## Accuracy of Death Certificates (A 6x3 Table)

The table below compiles data from six studies designed to investigate the accuracy of death certificates\footnote{Source:  Pagano M, Gauvreau K (2000) Principles of Biostatistics, 2nd Edition, Pacific Grove, CA: Duxbury, pp. 367-8.  The original citation is Kircher T, Nelson J, Burdo H (1985) The autopsy as a measure of accuracy of the death certificate.  NEJM, 313, 1263-1269.}.  5373 autopsies were compared to the causes of death listed on the certificates.  Of those, 3726 were confirmed to be accurate, 783 either lacked information or contained inaccuracies but did not require recoding of the underlying cause of death, and 864 were incorrect and required recoding.  Do the results across studies appear consistent?

Date of Study | [Confirmed] Accurate | [Inaccurate] No Change | [Incorrect] Recoding | Total
:-----------: | :------------------: | :--------------------: | :------------------: | :---:
1955-1965 | 2040 | 367 | 327 | 2734
1970 | 149 | 60 | 48 | 257
1970-1971 | 288 | 25 | 70 | 383
1975-1977 | 703 | 197 | 252 | 1152
1977-1978 | 425 | 62 | 88 | 575
1980 | 121 | 72 | 79 | 272
Total | 3726 | 783 | 864 | 5373

## The Pearson Chi-Square Test of Independence

We can assess the homogeneity of the confirmation results (columns) we observe in the table using a Pearson chi-squared test of independence.

- The null hypothesis is that the rows and columns are independent.
- The alternative hypothesis is that there is an association between the rows and the columns.

```{r death certificates a}
z <- matrix(c(2040,367,327,149,60,48,288,25,70,703,    
              197,252,425,62,88,121,72,79), byrow=TRUE, nrow=6)
rownames(z) <- c("1955-65", "1970", "1970-71", "1975-77", "1977-78",
                 "1980")
colnames(z) <- c("Confirmed", "Inaccurate", "Incorrect")
 
addmargins(z)
```

To see the potential heterogeneity across rows in these data, we should perhaps also look at the proportions of autopsies in each of the three accuracy categories for each study.

```{r death certificates b}
addmargins(round(100*prop.table(z,1),1),2) 
```

In three of the studies, approximately 3/4 of the results were confirmed. In the other three, 45%, 58% and 61% were confirmed. It looks like there's a fair amount of variation in results across studies. To see if this is true, formally, we run Pearson's chi-square test of independence, where the null hypothesis is that the rows and columns are independent, and the alternative hypothesis is that there is an association between the rows and the columns.

```{r death certificates c}
chisq.test(z)
```

The chi-square test statistic is 200 on 10 degrees of freedom, yielding p < 0.0001.

Autopsies are not performed at random; in fact, many are done because the cause of death listed on the certificate is uncertain.  What problems may arise if you attempt to use the results of these studies to make inference about the population as a whole?

