# Dehydration Recovery in Kids: A Small Study

The `hydrate` data describe the degree of recovery that takes place 90 minutes following treatment of moderate to severe dehydration, for 36 children diagnosed at a hospital's main pediatric clinic. 

Upon diagnosis and study entry, patients were treated with an electrolytic solution at one of seven `dose` levels (0, 0.5, 1.0, 1.5, 2.0, 2.5, 3.0 mEq/l) in a frozen, flavored, ice popsicle. The degree of rehydration was determined using a subjective scale based on physical examination and parental input, converted to a 0 to 100 point scale, representing the percent of recovery (`recov.score`). Each child's `age` (in years) and `weight` (in pounds) are also available. 

First, we'll check ranges (and for missing data) in the `hydrate` file. 

```{r show_hydrate_data}
hydrate
summary(hydrate)
```

There are no missing values, and all of the ranges make sense. There are no especially egregious problems to report.

## A Scatterplot Matrix

Next, we'll use a scatterplot matrix to summarize relationships between the outcome `recov.score` and the key predictor `dose` as well as the ancillary predictors `age` and `weight`, which are of less interest, but are expected to be related to our outcome. The one below uses the `ggpairs` function in the `GGally` package, as introduced in Part A of the Notes. We place the outcome in the bottom row, and the key predictor immediately above it, with `age` and `weight` in the top rows, using the `select` function within the ``ggpairs` call.

```{r scatterplot_matrix_hydrate_fig}
GGally::ggpairs(dplyr::select(hydrate, age, weight, dose, recov.score), 
                title = "Scatterplot Matrix for hydrate data")
```

What can we conclude here?

- It looks like `recov.score` has a moderately strong negative relationship with both `age` and `weight` (with correlations in each case around -0.5), but a positive relationship with `dose` (correlation = 0.36). 
- The distribution of `recov.score` looks to be pretty close to Normal. No potential predictors (`age`, `weight` and `dose`) show substantial non-Normality.
- `age` and `weight`, as we'd expect, show a very strong and positive linear relationship, with *r* = 0.94
- Neither `age` nor `weight` shows a meaningful relationship with `dose`. (*r* = 0.16)

## Are the recovery scores well described by a Normal model?

Next, we'll do a more thorough graphical summary of our outcome, recovery score, arranging the plots with the help of the `cowplot` package.

```{r recov.score_eda}
p1 <- ggplot(hydrate, aes(x = recov.score)) +
    geom_histogram(aes(y = ..density..),
                   bins = fd_bins(hydrate$recov.score), 
                   fill = '#440154', col = '#FDE725') +
    stat_function(fun = dnorm,
                  args = list(mean = mean(hydrate$recov.score), 
                              sd = sd(hydrate$recov.score)), 
                  lwd = 1.5, col = '#1FA187') +
    labs(title = "Histogram", x = "Recovery Score", y = "") +
    theme_bw() 

p2 <- ggplot(hydrate, aes(x = 1, y = recov.score)) + 
    geom_boxplot(fill = '#FDE725', notch = TRUE, 
                 col = '#440154', outlier.color = '#440154') +
    labs(title = "Boxplot", x = "", y = "") +
    theme_bw() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())  

p3 <- ggplot(hydrate, aes(sample = recov.score)) + 
    geom_qq(geom = "point", col = '#440154', size = 2) + 
    geom_abline(slope = qq_slope(hydrate$recov.score),
                intercept = qq_int(hydrate$recov.score),
                col = '#1FA187', size = 1.25) +
    labs(title = "Normal QQ", x = "", y = "") +
    theme_bw()

p <- cowplot::plot_grid(p1, p2, p3, align = "h", nrow = 1,
                   rel_widths = c(3, 1, 2))

title <- cowplot::ggdraw() + 
    cowplot::draw_label("Recovery Scores from 36 children in the hydrate study", 
               fontface = "bold")

cowplot::plot_grid(title, p, ncol = 1, rel_heights=c(0.1, 1))
```

I see no serious problems with assuming Normality for these recovery scores. Our outcome variable doesn't in any way *need* to follow a Normal distribution, but it's nice when it does, because summaries involving means and standard deviations make sense.




